Functional Pearl: the Proof Search Monad

### 2015-11-10 paper/main.tex
First pass on the reviewers' comments.

\usepackage{enumerate}
\usepackage{tabularx}
\usepackage{listings}
  \def\li{\lstinline}
  \lstset{
    language=caml,
    flexiblecolumns=false,
    showstringspaces=false,
    basicstyle=\ttfamily,
    framesep=5pt,
    numberstyle=\tiny,
    numbersep=5pt,
    escapeinside={(*}{*)},
    deletekeywords={value},
  }
\usepackage{minted}
  \newminted[ocaml]{ocaml}{mathescape,fontsize=\small}
\usepackage{mathpartir}
  % Improvements to Didier's mathpartir package: make all rule names and
  % references use textsc, and also make references to rules done using the \Rule
  % command clickable in a PDF.
  \let\TirName\textsc
  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
  \let\DefRule\RefTirName
  \let\Rule\DefTirName
  % \let\Rule\textsc

@@

@@

\begin{figure}
  \centering
\begin{ocaml}
let rec prove_equality (state: state) (goal: formula) (v1: var) (v2: var) =
  let open MOption in
  match P.find v1 state, P.find v2 state with
  | Flexible, Flexible
  | Flexible, Rigid ->
      let state = P.union v1 v2 state in
      prove goal begin
        premise (prove_equality state goal v1 v2) >>=
        qed R_Instantiate
      end
  (* ... *)
  | Rigid, Rigid ->
      if P.same v1 v2 state then
        axiom state goal R_Refl
      else
        fail

@@

@@

>>>>

\usepackage{enumerate}
\usepackage{tabularx}
\usepackage{listings}
  \def\li{\lstinline}
  \lstset{
    language=caml,
    flexiblecolumns=false,
    showstringspaces=false,
    basicstyle=\ttfamily,
    framesep=5pt,
    numberstyle=\tiny,
    numbersep=5pt,
    escapeinside={(*}{*)},
    deletekeywords={value},
  }
\usepackage{minted}
  \newminted[ocaml]{ocaml}{mathescape,fontsize=\small}
\usepackage{mathpartir}
  % Improvements to Didier's mathpartir package: make all rule names and
  % references use textsc, and also make references to rules done using the \Rule
  % command clickable in a PDF.
  \let\TirName\textsc
  % \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
  % \let\DefRule\RefTirName
  \let\Rule\DefTirName
  % \let\Rule\textsc

@@

This paper attempts to present, in a tutorial-style, the design of an OCaml
library. In order to facilitate the discussion, we focus on a very constrained
logic; later (\sref{extending}), we mention how to extend the library to
cover more use-cases. The original motivation for the library was to serve as a
core building block for the type-checker of Mezzo~\cite{protzenko-phd-14}. The
nature of the core, minimal logic that we are about to present is, of course,
inspired by typical nature of type-checking problems: it features equality,
quantifiers, and positive literals; \sref{extending} mentions how to extend it
with function symbols and variance (negative positions), as is typical for
type-checking problems.

@@

\begin{figure}
  \centering
\begin{ocaml}
let rec prove_equality (state: state) (goal: formula) (v1: var) (v2: var) =
  let open MOption in
  match P.find v1 state, P.find v2 state with
  | Flexible, Flexible
  | Flexible, Rigid ->
      let state = P.union v1 v2 state in
      prove goal begin
        (* Recursive call reduces to the [axiom ... R_Refl] case below. *)
        premise (prove_equality state goal v1 v2) >>=
        qed R_Instantiate
      end
  (* ... *)
  | Rigid, Rigid ->
      if P.same v1 v2 state then
        axiom state goal R_Refl
      else
        fail

@@

\section{Extending the library; extending the logic; limitations}
\label{sec:extending}

The approach advocated in the present paper works well for sequent-style
calculi; in the context of Mezzo, the logic is extended with the following extra
features:
\begin{itemize}
  \item function symbols, such as, but not limited to: ML arrow types ($\to$),
    type applications ($\mathsf{list}\ \alpha$), constructor applications
    ($\mathsf{Cons}\;\{ \mathsf{head}:\alpha;\ \mathsf{tail}:\mathsf{list}\
    \alpha\}$)
  \item positive and negative positions (also known as ``variance'' in
    type-checking lingo; this applies to function symbols, such as arrows or
    type constructors)
  \item higher-order quantification ($\forall (p: \mathsf{predicate}) \ldots$)
  \item affinity (where some hypothesis may be used at most once)
  \item framing, which bears some similarities with focusing.
\end{itemize}

This in turn requires the client code to keep track of more information, while
also adopting more sophisticated data structures. In particular, the client code
now carries, in addition to a substitution (\emph{a.k.a.} union-find), a
\emph{set} of available hypotheses, which flows left-to-right in the algorithmic
rules. Changing polarities changes the direction of the flow.

For the complete system (Mezzo), there are many more backtracking points. The
client code choses to backtrack on only a subset of them.

A limitation of this library is that it only works as long as every branch of
the exploration terminates; contrary to Kiselyov \emph{et al}.'s
library~\cite{oleg05}, we do not implement fair interleaving. One could
conceivably bound the depth of the search tree, but the exploration of the tree
remains sequential, not concurrent.

If one is willing to give up on modularity, stronger static guarantees can be
attained by making the \li+rule_name+ type more specific; namely, by encoding in
each constructor the number of premises required. The drawback is that the
library now has to be aware of the specific logic; in the current state of
things, the library is completely agnostic with regards to the client code's
particular logical system.

It is unclear how one could memoize the sub-computations; the state of the
client code (the union-find) may not change very much in the simplified logic
presented here; however, for a more evolved logic, it is likely that the client
state evolves monotonically, and makes memoizing difficult, as the result of
sub-calls depends very much on the state.

@@

\section*{Acknowledgments}

I wish to thank Gabriel Scherer who helped me sketch the initial version of the
library; François Pottier for motivating me enough that I would want to write
about it, as well as providing the persistent union-find implementation; and
Anonymous Review \#2 for a very high quality review with numerous excellent
suggestions.
